---
title: "Get data from BIEN"
author: "Brunno Oliveira"
date: as.character(Sys.Date())
output:
  pdf_document: default
  html_document: default
---

# Setup
```{r message=F,warning=FALSE, results='hide'}

rm(list=ls())
gc()

list.of.packages <- c("BIEN",
                      "TNRS",
                      "data.table", # for loading large tables
                      "dplyr", # data organization
                      "ape", #Package for working with phylogenies in R
                      "maps","rgdal","raster","leaflet", # Useful for making quick maps of occurrences
                      "sp", "mapview","plainview","sf","fasterize", # More mapping
                      "foreach", "pbapply", "doParallel") # parallel processing

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

sapply(list.of.packages, require, character.only = TRUE)

```

# Download BIEN traits
Dowloaded 11/08/2021

We downloaded data for the 16 traits from BIEN. These are among the traits with highest completeness.
* leaf area (LA)
* leaf area per leaf dry mass (SLA)
* leaf carbon content per leaf dry mass (C)
* leaf dry mass (LDM)
* leaf fresh mass (LFM)
* leaf nitrogen content per leaf dry mass (N)
* leaf phosphorus content per leaf dry mass (P)
* leaf photosynthetic rate per leaf area (LPhoto)
* leaf stomatal conductance per leaf area (LStom)
* leaf thickness (Thick)
* longest whole plant longevity / maximum whole plant longevity (Lon)
* whole plant height / maximum whole plant height (Height)
* root dry mass (RootMass)
* seed mass (SeedMass)
* stem dry mass (StemMass)
* stem wood density (WoodDens)

Also downloaded this extra trait for use in analyses of dispersion distance
* Plant growth form

```{r traits, eval=FALSE, message=FALSE, warning=FALSE}
# traits_available <- BIEN_trait_list()
# 
# # Download all measurements of the trait(s) specified
# traits2download <- c(LA = "leaf area",
#                      SLA = "leaf area per leaf dry mass",
#                      LeafCarb = "leaf carbon content per leaf dry mass",
#                      LeafDryMass = "leaf dry mass",
#                      Tickness = "leaf thickness",
#                      LeafFreshMass = "leaf fresh mass",
#                      N = "leaf nitrogen content per leaf dry mass",
#                      P = "leaf phosphorus content per leaf dry mass",
#                      LeafPhotoRateArea = "leaf photosynthetic rate per leaf area",
#                      LeafStomatCondArea = "leaf stomatal conductance per leaf area",
#                      Long = "longest whole plant longevity",
#                      LongMax = "maximum whole plant longevity",
#                      HeightMax = "maximum whole plant height",
#                      RootDryMass = "root dry mass",
#                      Height = "whole plant height",
#                      SeedMass = "seed mass",
#                      SDM = "stem dry mass",
#                      WoodDens = "stem wood density")
# 
# # To calc
# # Leaf nitrogen/phosphorus (N/P) ratio
# # Leaf area per leaf fresh mass (SLA based on leaf fresh mass)
# 
# BIEN_traits <- pblapply(traits2download, function(x) BIEN_trait_trait(trait = x))
# # Filter important columns
# BIEN_traits <- lapply(BIEN_traits, function(x) x[,1:3])
# 
# # Fix height and longevity
# ## 1) height
# Height <- rbind(BIEN_traits$Height, BIEN_traits$HeightMax)
# Height <- Height %>%
#   group_by(scrubbed_species_binomial) %>%
#   summarise(trait_value = max(as.numeric(trait_value), na.rm = T))
# Height$trait_name = "maximum whole plant height"
# BIEN_traits$Height <- Height
# BIEN_traits <- BIEN_traits[-which(names(BIEN_traits) == "HeightMax")]
# 
# ## 2) longevity
# Long <- rbind(BIEN_traits$Long, BIEN_traits$LongMax)
# Long <- Long %>%
#   group_by(scrubbed_species_binomial) %>%
#   summarise(trait_value = max(as.numeric(trait_value), na.rm = T))
# Long$trait_name = "longest whole plant longevity"
# BIEN_traits$Long <- Long
# BIEN_traits <- BIEN_traits[-which(names(BIEN_traits) == "LongMax")]
# 
# # save continuous traits
# BIEN_traits_cont <- do.call(rbind, BIEN_traits)
# 
# saveRDS(BIEN_traits_cont, "Data/BIEN_traits_cont.RDS")

# #############################
# 
# # Download categorical traits
# 
# # PS.: I did not download "whole plant dispersal syndrome" because they classify this trait as eigther "abiotic" or "biotic". TRY database has a much better/detailed classification
# 
# # The other important traits is then "whole plant growth form"
# 
# GF <- BIEN_trait_trait(trait = "whole plant growth form")[,1:3]
# unique(GF$trait_value)
# 
# saveRDS(GF, "Data/BIEN_traits_growthform.RDS")

# Continuous traits
BIEN_traits_cont <- readRDS("Data/BIEN_traits_cont.RDS")
colnames(BIEN_traits_cont) <- c("sps", "trait", "value")
# trait value as numeric
BIEN_traits_cont$value <- as.numeric(BIEN_traits_cont$value)
# omit NAs
BIEN_traits_cont <- na.omit(BIEN_traits_cont)

# Fix trait names
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf area")] <- "LA"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf area per leaf dry mass")] <- "SLA"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf carbon content per leaf dry mass")] <- "C"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf dry mass")] <- "LDM"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf fresh mass")] <- "LFM"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf nitrogen content per leaf dry mass")] <- "N"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf phosphorus content per leaf dry mass")] <- "P"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf photosynthetic rate per leaf area")] <- "LPho"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf stomatal conductance per leaf area")] <- "LStom"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="leaf thickness")] <- "LThick"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="longest whole plant longevity")] <- "Lon"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="maximum whole plant height")] <- "Height"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="root dry mass")] <- "RootMass"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="seed mass")] <- "SeedMass"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="stem dry mass")] <- "StemMass"
BIEN_traits_cont$trait[which(BIEN_traits_cont$trait=="stem wood density")] <- "WoodDens"

sort(unique(BIEN_traits_cont$trait))

BIEN_traits_cont$value <- as.numeric(BIEN_traits_cont$value)
BIEN_traits_cont <- BIEN_traits_cont[-which(is.infinite(BIEN_traits_cont$value)),]
BIEN_traits_cont <- na.omit(BIEN_traits_cont)

# Categorical traits
BIEN_GF <- readRDS("Data/BIEN_traits_growthform.RDS")
colnames(BIEN_GF) <- c("sps", "trait", "value")
# fix sp names
BIEN_GF$sps <- gsub(" ","_",BIEN_GF$sps)
# omit NAs
BIEN_GF <- na.omit(BIEN_GF)
BIEN_GF$trait <- "GF"

```

# Load TRY data

We downloaded data for the 9 most complete traits from TRY: 

* Leaf area
* Leaf area per leaf dry mass
* Leaf carbon (C) content per leaf dry mass
* Leaf carbon/nitrogen (C/N) ratio
* Leaf dry mass (single leaf)
* Leaf dry mass per leaf fresh mass (leaf dry matter content, LDMC)
* Leaf fresh mass
* Leaf nitrogen (N) content per leaf dry mass
* Leaf nitrogen/phosphorus (N/P) ratio
* Leaf phosphorus (P) content per leaf dry mass
* Leaf photosynthesis rate per leaf area
* Leaf potassium (K) content per leaf dry mass
* Leaf thickness
* Leaf transpiration rate per leaf area
* Leaf water content per leaf dry mass (at saturation)
* Plant biomass and allometry: Leaf dry mass per plant dry mass (leaf weight ratio, LWR)
* Plant biomass and allometry: Shoot dry mass (plant aboveground dry mass) per plant
* Plant biomass and allometry: Whole plant aboveground mass
* Plant height vegetative
* Plant lifespan (longevity)
* Root rooting depth
* Seed (seedbank) longevity
* Seed dry mass
* Seed germination rate (germination efficiency)
* Stem specific density (SSD) or wood density (stem dry mass per stem fresh volume)
* Stomata conductance per leaf area

Also downloaded these extra traits for use in analyses of dispersion distance
* Dispersal syndrome
* Plant growth form
* Seed terminal velocity

```{r}

TRY_traits <- fread("C:/TRY_requests/17401_09112021075727/17401.txt", header = T, sep = "\t", quote="")

sppnames <- unique(TRY_traits$AccSpeciesName)
sppnames <- data.frame(sps = sppnames,
                       New = sppnames)
# Remove species from which there is only genus and no species level info
rem <- sapply(sppnames$New, function(x) length(strsplit(x," ")[[1]])) 
sppnames <- sppnames[-which(rem==1),]
# Get only genus and species names, remove subspecies (BIEN's data does not have subspecies, so species level is fine here!)
sppnames$New <- sapply(sppnames$New, 
                       function(x){ 
                         split. <- strsplit(x," ")[[1]]
                         paste(split.[1],split.[2])}) 

sppnames <- sppnames[-duplicated(sppnames$New),]

# Clean TRY data
# get columns of interest
TRY_traits_cont <- TRY_traits[,c(7,11,21)]
TRY_traits_cat <- TRY_traits[,c(7,11,15)]

colnames(TRY_traits_cont) <- c("sps", "trait", "value")
colnames(TRY_traits_cat) <- c("sps", "trait", "value")

TRY_traits_cont <- merge(TRY_traits_cont, sppnames,by = "sps", all.x = T)
TRY_traits_cat <- merge(TRY_traits_cat, sppnames,by = "sps", all.x = T)

TRY_traits_cont$sps <- TRY_traits_cont$New
TRY_traits_cat$sps <- TRY_traits_cat$New

TRY_traits_cont <- TRY_traits_cont[,c(1:3)]
TRY_traits_cat <- TRY_traits_cat[,c(1:3)]

TRY_traits_cont <- na.omit(TRY_traits_cont)
TRY_traits_cat <- na.omit(TRY_traits_cat)

TRY_traits_cont <- TRY_traits_cont[-which(TRY_traits_cont$trait==""),]
TRY_traits_cat <- TRY_traits_cat[-which(TRY_traits_cat$trait==""),]

# Traits of interest
interest <- c("Dispersal syndrome", "Plant growth form")
TRY_traits_cat <- TRY_traits_cat[which(TRY_traits_cat$trait %in% interest),]

TRY_GF <- TRY_traits_cat[which(TRY_traits_cat$trait == "Plant growth form"),]
TRY_GF$trait <- "GF"

Disp <- TRY_traits_cat[which(TRY_traits_cat$trait == "Dispersal syndrome"),]
Disp$trait <- "Disp"

rm(TRY_traits,TRY_traits_cat)

```

# Combine trait types
```{r}

traits_cont <- rbind(TRY_traits_cont, BIEN_traits_cont)

GF <- rbind(BIEN_GF,TRY_GF)
GF <- na.omit(GF)

rm(TRY_traits_cont, BIEN_traits_cont, BIEN_GF, TRY_GF)

```

## Fix continuous traits
```{r}

# Leaf area
posit <- grepl("Leaf area \\(in case of ",traits_cont$trait)
traits_cont$trait[posit] <- "Leaf area"
# Fix Leaf area per leaf dry mass
posit <- grepl("Leaf area per leaf dry mass",traits_cont$trait)
traits_cont$trait[posit] <- "Leaf area per leaf dry mass"

# Check
unique(traits_cont$trait)
length(unique(traits_cont$sps))

traits_cont$value <- as.numeric(traits_cont$value)
traits_cont <- na.omit(traits_cont)

traits_cont$trait[which(traits_cont$trait=="Leaf area")] <- "LA"
traits_cont$trait[which(traits_cont$trait=="Leaf area per leaf dry mass")] <- "LADM"
traits_cont$trait[which(traits_cont$trait=="Leaf carbon (C) content per leaf dry mass")] <- "C"
traits_cont$trait[which(traits_cont$trait=="Leaf carbon/nitrogen (C/N) ratio")] <- "CN"
traits_cont$trait[which(traits_cont$trait=="Leaf dry mass (single leaf)")] <- "LDM"
traits_cont$trait[which(traits_cont$trait=="Leaf dry mass per leaf fresh mass (leaf dry matter content, LDMC)")] <- "LDMC"
traits_cont$trait[which(traits_cont$trait=="Leaf fresh mass")] <- "LFM"
traits_cont$trait[which(traits_cont$trait=="Leaf nitrogen (N) content per leaf dry mass")] <- "N"
traits_cont$trait[which(traits_cont$trait=="Leaf nitrogen/phosphorus (N/P) ratio")] <- "NP"
traits_cont$trait[which(traits_cont$trait=="Leaf phosphorus (P) content per leaf dry mass")] <- "P"
traits_cont$trait[which(traits_cont$trait=="Leaf photosynthesis rate per leaf area")] <- "LPho"
traits_cont$trait[which(traits_cont$trait=="Leaf potassium (K) content per leaf dry mass")] <- "K"
traits_cont$trait[which(traits_cont$trait=="Leaf thickness")] <- "LThick"
traits_cont$trait[which(traits_cont$trait=="Leaf transpiration rate per leaf area")] <- "LTransp"
traits_cont$trait[which(traits_cont$trait=="Leaf water content per leaf dry mass (at saturation)")] <- "LWater"
traits_cont$trait[which(traits_cont$trait=="Plant biomass and allometry: Leaf dry mass per plant dry mass (leaf weight ratio, LWR)")] <- "LWR"
traits_cont$trait[which(traits_cont$trait=="Plant biomass and allometry: Shoot dry mass (plant aboveground dry mass) per plant")] <- "ShootDM"
traits_cont$trait[which(traits_cont$trait=="Plant biomass and allometry: Whole plant aboveground mass")] <- "AbovMass"
traits_cont$trait[which(traits_cont$trait=="Plant height vegetative")] <- "Height"
traits_cont$trait[which(traits_cont$trait=="Plant lifespan (longevity)")] <- "Lon"
traits_cont$trait[which(traits_cont$trait=="Root rooting depth")] <- "RootDepth"
traits_cont$trait[which(traits_cont$trait=="Seed (seedbank) longevity")] <- "SeedBankLon"
traits_cont$trait[which(traits_cont$trait=="Seed dry mass")] <- "SeedMass"
traits_cont$trait[which(traits_cont$trait=="Seed germination rate (germination efficiency)")] <- "SeedGermRate"
traits_cont$trait[which(traits_cont$trait=="Stem specific density (SSD) or wood density (stem dry mass per stem fresh volume)")] <- "WoodDens"
traits_cont$trait[which(traits_cont$trait=="Stomata conductance per leaf area")] <- "LStom"
traits_cont$trait[which(traits_cont$trait=="Seed terminal velocity")] <- "SeedVel"

unique(traits_cont$trait)

```


# Resolve species names using TNRS
Resolve names based on the TAXONOMIC NAME RESOLUTION SERVICE V5.0
```{r}

Ori <- unique(c(traits_cont$sps,
                Disp$sps,
                GF$sps))

# Remove plant names containing 'sp.', 'subsp.', 'subgen.', 'sect.' and 'cf.'
if(any(grep('sp\\.',Ori))){ Ori <- Ori[-grep('sp\\.',Ori)] }
if(any(grep('subgen\\.',Ori))){ Ori <- Ori[-grep('subgen\\.',Ori)] }
if(any(grep('sect\\.',Ori))){ Ori <- Ori[-grep('sect\\.',Ori)] }
if(any(grep('cf\\.',Ori))){ Ori <- Ori[-grep('cf\\.',Ori)]  }

Ori <- gsub("_"," ",Ori)

length(unique(Ori)) # 146364

# Run in parallel for faster results
spchunks <- split(Ori, sort(rep_len(1:160, length(Ori))))
# N species in each chunck
# head(lapply(spchunks, length))

cl <- makeCluster(detectCores())
clusterExport(cl, c("spchunks", "TNRS"))

TNRS_names <- pblapply(spchunks, function(i) {
  test <- TNRS(taxonomic_names = i)
  while(length(test)==0){ # prevent error
    test <- TNRS(taxonomic_names = i)
  }
  return(test)
  }
  , cl = cl
)

stopCluster(cl)
# took 10m 30s

head(lapply(TNRS_names, dim))

# check if any chunk fail
cfail <- sapply(TNRS_names, function(x) {
  if(length(x)==0){ 1 }
  else{ 0 }
})
# N chunks that fail
length(which(cfail==1))

TNRS_names <- do.call("rbind", TNRS_names)

# remove species not found
TNRS_names <- TNRS_names[-which(TNRS_names$Overall_score==""),]

# remove bryophytes
taxonomy <- BIEN::BIEN_taxonomy_species(gsub("_"," ",TNRS_names$Name_matched))
length(unique(taxonomy$scrubbed_species_binomial))
length(unique(TNRS_names$Name_matched))
taxonomy2 <- taxonomy[-which(duplicated(taxonomy$scrubbed_species_binomial)),]
TNRS_names_ <- merge(taxonomy2[,c("scrubbed_species_binomial","higher_plant_group")], TNRS_names, 
                     by.x = "scrubbed_species_binomial", by.y = "Name_matched",
                     all.x = T)
# remove bryophytes
TNRS_names_ <- TNRS_names_[-which(TNRS_names_$higher_plant_group=="bryophytes"),]

my_sp_list <- unique(TNRS_names_$scrubbed_species_binomial)
length(my_sp_list)
# N species = 139345

# Fix names
traits_cont <- merge(traits_cont, 
                     TNRS_names_[,c("Name_submitted","scrubbed_species_binomial")], 
                     by.x = "sps", by.y = "Name_submitted",
                     all.x = T,
                     allow.cartesian=TRUE) # necessary in this case due to large matrix
traits_cont$sps <- traits_cont$scrubbed_species_binomial
traits_cont <- traits_cont[,1:3]
traits_cont <- na.omit(traits_cont)

GF <- merge(GF, 
            TNRS_names_[,c("Name_submitted","scrubbed_species_binomial")], 
            by.x = "sps", by.y = "Name_submitted",
            all.x = T)
GF$sps <- GF$scrubbed_species_binomial
GF <- GF[,1:3]
GF <- na.omit(GF)

Disp <- merge(Disp, 
              TNRS_names_[,c("Name_submitted","scrubbed_species_binomial")], 
              by.x = "sps", by.y = "Name_submitted",
              all.x = T)
Disp$sps <- Disp$scrubbed_species_binomial
Disp <- Disp[,1:3]
Disp <- na.omit(Disp)

```

# Remove species not in the the sp list
```{r}

mysplist <- readRDS("R/3_SDM/Cluster/plants_list.RDS")

traits_cont <- traits_cont[which(traits_cont$sps %in% mysplist),]
GF <- GF[which(GF$sps %in% mysplist),]
Disp <- Disp[which(Disp$sps %in% mysplist),]

```


# Fix traits categorical
## Growth form
```{r fix, eval=FALSE, message=FALSE, warning=FALSE}

GF$value[which(grepl("tree",GF$value,ignore.case = T))] <- "Tree"
GF$value[which(grepl("palm",GF$value,ignore.case = T))] <- "Tree"
GF$value[which(grepl("pygmy",GF$value,ignore.case = T))] <- "Tree"

GF$value[which(grepl("shrub",GF$value,ignore.case = T))] <- "Shrub"
GF$value[which(grepl("mallee",GF$value,ignore.case = T))] <- "Shrub"
GF$value[which(grepl("suffrut",GF$value,ignore.case = T))] <- "Shrub"
GF$value[which(grepl("Bush",GF$value,ignore.case = T))] <- "Shrub"
GF$value[which(grepl("shub",GF$value,ignore.case = T))] <- "Shrub"
GF$value[which(grepl("shurb",GF$value,ignore.case = T))] <- "Shrub"
GF$value[which(grepl("prostrate",GF$value,ignore.case = T))] <- "Shrub"

GF$value[which(grepl("herb",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("hierb",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("forb",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("hepatic",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("herg",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("wildflower",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("marsh",GF$value,ignore.case = T))] <- "Herb"
GF$value[which(grepl("crucifer",GF$value,ignore.case = T))] <- "Herb"

GF$value[which(grepl("gras",GF$value,ignore.case = T))] <- "Grass"
GF$value[which(grepl("bamboo",GF$value,ignore.case = T))] <- "Grass"
GF$value[which(grepl("sedge",GF$value,ignore.case = T))] <- "Grass"
GF$value[which(grepl("gramin",GF$value,ignore.case = T))] <- "Grass"
GF$value[which(grepl("bunchgrass",GF$value,ignore.case = T))] <- "Grass"
GF$value[which(grepl("tussock",GF$value,ignore.case = T))] <- "Grass"
GF$value[which(grepl("cyperoid",GF$value,ignore.case = T))] <- "Grass"

GF$value[which(grepl("succul",GF$value,ignore.case = T))] <- "Succulent"
GF$value[which(grepl("aloe",GF$value,ignore.case = T))] <- "Succulent"
GF$value[which(grepl("cact",GF$value,ignore.case = T))] <- "Succulent"

  GF$value[which(grepl("creeper",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("vine",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("epiphy",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("climb",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("fern",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("liana",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("lianna",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("twin",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("scrambler",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("trail",GF$value,ignore.case = T))] <- "Epiphyte"
GF$value[which(grepl("scandent",GF$value,ignore.case = T))] <- "Epiphyte"

GF$value[which(grepl("moss",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("lycop",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("cushion",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("liverwort",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("brophyte",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("bryoid",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("acrocarpous",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("pleurocarpus",GF$value,ignore.case = T))] <- "Moss-like"

GF$value[which(grepl("brophyte",GF$value,ignore.case = T))] <- "Moss-like"
GF$value[which(grepl("bryoid",GF$value,ignore.case = T))] <- "Moss-like"

GF$value[which(grepl("Reed",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("hidro",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("hydro",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("aquatic",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("phaerophyt",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("Submerge",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("Emergent",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("Float",GF$value,ignore.case = T))] <- "Aquatic"
GF$value[which(grepl("Macrophyte",GF$value,ignore.case = T))] <- "Aquatic"

# View(unique(GF$value))

# remove all GFs not in this list
cases <- c("Tree", "Palm", "Shrub", "Herb", "Grass", "Succulent", "Epiphyte", "Moss-like", "Aquatic")

GF <- GF[which(GF$value %in% cases),]

# View(unique(GF$value))

# Remove Moss-like
GF <- GF[-which(GF$value=="Moss-like"),]
# Remove aquatic
GF <- GF[-which(GF$value=="Aquatic"),]

# One trait category per sps
## N obs per trait
GF_n <- GF %>%
  group_by(sps, value) %>%
  summarise(nobs = length(value)) # n obs per value

## Protocol for selecting a single GF value for species classified in multiple categories of GF
## 1) If n obs is equal among GF values, use priority scheme
mn = length(cases)
priority <- 1:mn
names(priority) <- cases[mn:1]
## 2) Use the most commonly used GF

## splist with more then 1 GF per sps
sp.listgo <- GF_n$sps[which(duplicated(GF_n$sps))]
sp.listgo <- unique(sp.listgo)

# GF with one obs per sps
GF1 <- GF_n[-which(GF_n$sps %in% sp.listgo),]

# select one GF per sps
GF2 <- pblapply(sp.listgo, function(x){
  tmp <- GF_n[which(GF_n$sps == x),]
  if(dim(tmp)[1]>1){
    # condition 1
    if(var(tmp$nobs) == 0){
      sel <- priority[tmp$value]
      sel <- order(sel,decreasing = T)[1]
      tmp <- tmp[sel,]
    }
    else{
      # condition 2
      sel <- order(tmp$nobs,decreasing = T)[1]
      tmp <- tmp[sel,]
    }
  }
  return(data.frame(tmp))
})
GF2 <- do.call(rbind,GF2)

# Group
GF <- rbind(GF1, GF2)
GF <- data.frame(sps = GF$sps, trait = "GF", value = GF$value)

rm(GF1, GF2, BIEN_GF, TRY_GF, GF_n)

```

## Disp
Following: https://nutnet.org/sites/default/files/Leda_dispersability%20traits.pdf
```{r}

# View(unique(Disp$value))

# Autochor
## Autochor: Self dispersal
## Ballochor: Explosive mechanisms
## Blastochor: Autonomous placement of seeds or daughter plant away from mother plant
Disp$value[which(grepl("autochor",Disp$value,ignore.case = T))] <- "Autochor"
Disp$value[which(grepl("Ballochor",Disp$value,ignore.case = T))] <- "Autochor"
Disp$value[which(grepl("Blastochor",Disp$value,ignore.case = T))] <- "Autochor"
Disp$value[which(grepl("drop",Disp$value,ignore.case = T))] <- "Autochor"
Disp$value[which(grepl("explo",Disp$value,ignore.case = T))] <- "Autochor"
Disp$value[which(grepl("Barochory",Disp$value,ignore.case = T))] <- "Autochor" # seed drop
Disp$value[which(grepl("herpochor",Disp$value,ignore.case = T))] <- "Autochor" 
Disp$value[which(grepl("gravity",Disp$value,ignore.case = T))] <- "Autochor" 

# Chamaechor: Tumbleweeds; dispersal unit rolling over the soil surface caused by wind
Disp$value[which(grepl("Chamaechor",Disp$value,ignore.case = T))] <- "Chamaechor"
Disp$value[which(grepl("Tumble",Disp$value,ignore.case = T))] <- "Chamaechor"
Disp$value[which(grepl("rolling",Disp$value,ignore.case = T))] <- "Chamaechor"
Disp$value[which(grepl("tumbling",Disp$value,ignore.case = T))] <- "Chamaechor"

# Ombrochor: ‘Raindrop-ballists’: raindrops triggering ballistic seed dispersal 
Disp$value[which(grepl("Ombrochor",Disp$value,ignore.case = T))] <- "Ombrochor"
Disp$value[which(grepl("ballis",Disp$value,ignore.case = T))] <- "Ombrochor"
Disp$value[which(grepl("Rain",Disp$value,ignore.case = T))] <- "Ombrochor"

# Anemochor:  Dispersal by wind (Note: flyers only, no tumbleweeds or wind-ballistics)
Disp$value[which(grepl("anemocho",Disp$value,ignore.case = T))] <- "Anemochor"
Disp$value[which(grepl("anemo",Disp$value,ignore.case = T))] <- "Anemochor"
Disp$value[which(grepl("wind",Disp$value,ignore.case = T))] <- "Anemochor"
Disp$value[which(grepl("meteorochor",Disp$value,ignore.case = T))] <- "Anemochor"
Disp$value[which(grepl("wing",Disp$value,ignore.case = T))] <- "Anemochor" # seeds with wings
Disp$value[which(grepl("unassisted",Disp$value,ignore.case = T))] <- "Anemochor" # seeds with wings
Disp$value[which(grepl("boleochor",Disp$value,ignore.case = T))] <- "Anemochor"
Disp$value[which(grepl("fly",Disp$value,ignore.case = T))] <- "Anemochor"

#  Zoochor
## Dysochor: Dispersal by scatter-hoarding animals
## Endozoochor: Dispersal after digestion
## Epizoochor: Adhesive dispersal
Disp$value[which(grepl("Zoochor",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Dysochor",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Endozoochor",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Epizoochor",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("animal",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("adhesion",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("bird",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("mammal",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("zoo",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("dysochor",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("pig",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("sheep",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("cattle",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("eaten",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("bear",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("monkey",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("ant",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("myrme",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("vertebrate",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("crab",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("crustacea",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("equus",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("capra",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("ruminant",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("cow",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("goat",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("gazell",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("elephant",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("livestock",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("mouse",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("squirrel",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("insect",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("ornitho",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("deer",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("dog",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("boar",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("rabbit",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Adhesi",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("hooks",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("spine",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("donkey",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("horse",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("bats",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("reptile",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("iguana",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("alligator",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Macaque",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("turtle",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Toucans",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("buffalo",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Mamal",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Macaca",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("rodent",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("peccary",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Canis",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("mule",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("sparrow",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Lizard",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("gorilla",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("chimpanzee",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("beetle",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("invertebrate",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("primate",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("pigeon",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("finche",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("tanager",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("flycatcher",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("cotinga",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("herbivore",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("possum",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Ursus",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Canis",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("shrew",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl("Giraffe",Disp$value,ignore.case = T))] <- "Zoochor"
Disp$value[which(grepl(" bat",Disp$value,ignore.case = T))] <- "Zoochor"

# Hemerochor
## Agochor: Unintended dispersal by man
## Ethelocho: Dispersal by trading of plants or seeds
## Hemerochor: Dispersal by man
## Speirochor: Dispersal with seeds of agricultural species
Disp$value[which(grepl("Hemerochor",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("ethelochor",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("man",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("trade",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("agricul",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("commerce",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("clothe",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("footwear",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("human",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("speirochor",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("agochor",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("vehicle",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("contamination",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("machinery",Disp$value,ignore.case = T))] <- "Hemerochor"
Disp$value[which(grepl("transport",Disp$value,ignore.case = T))] <- "Hemerochor"

# Nautochor: Dispersal by surface currents of water
Disp$value[which(grepl("Nautochor",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("Water",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("fish",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("float",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("current",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("hidro",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("hydro",Disp$value,ignore.case = T))] <- "Nautochor"
Disp$value[which(grepl("bythisochor",Disp$value,ignore.case = T))] <- "Nautochor"

# View(unique(Disp$value))

# remove all GFs not in this list
cases <- c("Autochor", "Chamaechor", "Ombrochor", "Anemochor", "Zoochor", "Hemerochor", "Nautochor")

Disp <- Disp[which(Disp$value %in% cases),]

Disp <- Disp

# One trait category per sps
## N obs per trait
Disp_n <- Disp %>%
  group_by(sps, value) %>%
  summarise(nobs = length(value)) # n obs per value

## Protocol for selecting a single Disp value for species classified in multiple categories of Disp
## 1) If n obs is equal among Disp values, use priority scheme
mn = length(cases)
priority <- 1:mn
names(priority) <- cases[mn:1]
## 2) Use the most commonly used Disp

## splist with more then 1 Disp per sps
sp.listgo <- Disp_n$sps[which(duplicated(Disp_n$sps))]
sp.listgo <- unique(sp.listgo)

# Disp with one obs per sps
Disp1 <- Disp_n[-which(Disp_n$sps %in% sp.listgo),]

# select one Disp per sps
Disp2 <- pblapply(sp.listgo, function(x){
  tmp <- Disp_n[which(Disp_n$sps == x),]
  if(dim(tmp)[1]>1){
    # condition 1
    if(var(tmp$nobs) == 0){
      sel <- priority[tmp$value]
      sel <- order(sel,decreasing = T)[1]
      tmp <- tmp[sel,]
    }
    else{
      # condition 2
      sel <- order(tmp$nobs,decreasing = T)[1]
      tmp <- tmp[sel,]
    }
  }
  return(data.frame(tmp))
})
Disp2 <- do.call(rbind,Disp2)

# Group
Disp <- rbind(Disp1, Disp2)
Disp <- data.frame(sps = Disp$sps, trait = "Disp", value = Disp$value)

# rm(Disp1, Disp2, Disp_n, Disp)

```

# save
```{r}

traits_cat <- rbind(GF, Disp)

saveRDS(traits_cont, "Data/plants_traits_cont.RDS")
saveRDS(traits_cat, "Data/plants_traits_cat.RDS")

```



