---
title: "Species redistribution - Get predictor variables for use in SDMs"
author: "Brunno F Oliveira"
date: "9/28/2021"
output:   
  html_document:
    toc: true
    toc_float: true
    code_folding: hide 
---

# Setup
```{r packages, message=FALSE, include=FALSE}

rm(list=ls())
gc()

list.of.packages <- c("raster","spdep","rgdal","rworldmap","maptools","gdalUtils","sf","vegan","adespatial","geodata","envirem",
                      "dismo","rJava",
                      "gridExtra","viridis","ggplot2","cowplot",
                      "reshape2","tidyr","tidyverse",
                      "RCurl","curl",
                      "soilDB",
                      "foreach","doParallel","pbapply","bigdist")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

sapply(list.of.packages, require, character.only = TRUE)

```

# Source
```{r}
source("R/ETsolradRasters.r")
```

# Download predictor variables

## Bioclimatic (Worldclim)
Worldclim ver. 1.4 (Hijmans et al. 2005)

Worldclim variables code:
BIO1 = Annual Mean Temperature
BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
BIO3 = Isothermality (BIO2/BIO7) (* 100)
BIO4 = Temperature Seasonality (standard deviation *100)
BIO5 = Max Temperature of Warmest Month
BIO6 = Min Temperature of Coldest Month
BIO7 = Temperature Annual Range (BIO5-BIO6)
BIO8 = Mean Temperature of Wettest Quarter
BIO9 = Mean Temperature of Driest Quarter
BIO10 = Mean Temperature of Warmest Quarter
BIO11 = Mean Temperature of Coldest Quarter
BIO12 = Annual Precipitation
BIO13 = Precipitation of Wettest Month
BIO14 = Precipitation of Driest Month
BIO15 = Precipitation Seasonality (Coefficient of Variation)
BIO16 = Precipitation of Wettest Quarter
BIO17 = Precipitation of Driest Quarter
BIO18 = Precipitation of Warmest Quarter
BIO19 = Precipitation of Coldest Quarter

We downloaded data for bioclimatic variables calculated using the new Coupled Model Intercomparison Project Phase 6 (CMIP6) with projections for "2061-2080".
- 3 global circulation models (GCM): MIROC6, MRI-ESM2-0, and IPSL-CM6A-LR.
- 2 Shared Socioeconomic Pathways (SSPs): ssp126 to approximate climate change consistent with meeting the Paris Agreement target of 2.0°C global mean temperature change, and ssp585 to approximate business-as-usual, no action on climate change.
Citation: Eyring, V., Bony, S., Meehl, G. A., Senior, C. A., Stevens, B., Stouffer, R. J., and Taylor, K. E.: Overview of the Coupled Model Intercomparison Project Phase 6 (CMIP6) experimental design and organization, Geosci. Model Dev., 9, 1937-1958, doi:10.5194/gmd-9-1937-2016, 2016.
```{r clim}
# present
if(!dir.exists("R/3_SDM/predictors/")){
  dir.create("R/3_SDM/predictors/")
}

climatelayers <- worldclim_global(var = "bio", 
                                  res = 5, 
                                  path = "R/3_SDM/predictors") # 5 arc min ~ 10km ~ 0.08333 Dg

# future
if(!dir.exists("R/3_SDM/predictors/future")){
  dir.create("R/3_SDM/predictors/future")
}

gcms <- c("MIROC6", "MRI-ESM2-0", "IPSL-CM6A-LR")
ssps <- c(126, 585)
varstodown <- expand.grid(gcms, ssps)

climatelayers_fut <- lapply(1:nrow(varstodown),
                            function(x){
                              cmip6_world(model = varstodown[x,1],
                                          ssp = varstodown[x,2], 
                                          time = "2061-2080", 
                                          var = "bioc", 
                                          res = 5, 
                                          path = "R/3_SDM/predictors/future")})

```

## Aridity
Global aridity index from Trabucco and Zomer (2019):
Trabucco, A. and Zomer, R. 2019. Global aridity index and potential evapotranspiration (ET0) climate database v2. figshare. Dataset. – <https://doi.org/10.6084/m9.figshare.7504448.v3>
```{r ai}

AI <- raster("D:/GIS/#Environment/Global PET and Aridity Index/global-ai_et0/ai_et0/ai_et0.tif")
AI <- AI * 0.0001 # Global Aridity Index_ET0 values need to be multiplied by0.0001 to retrieve the values in the correct units.
AI <- resample(AI,raster(climatelayers[[1]]))

writeRaster(AI, "R/3_SDM/predictors/aridity.tif",format = "GTiff", overwrite=TRUE)

```

# Additional variables
```{r}

# Altitude
elev <- elevation_global(res = 5, 
                         path = "R/3_SDM/predictors")
elev <- raster(elev)

# calculate aridity for the present for comparizon with metri from Trabucco and Zomer (2019)

# tavg
if(!dir.exists("R/3_SDM/predictors/tavg")){
  dir.create("R/3_SDM/predictors/tavg")
}
tavg <- worldclim_global(var = "tavg", 
                         res = 5, 
                         path = "R/3_SDM/predictors/tavg") # 5 arc min ~ 10km ~ 0.08333 Dg
tavg <- stack(tavg)
# tmin
if(!dir.exists("R/3_SDM/predictors/tmin")){
  dir.create("R/3_SDM/predictors/tmin")
}
tmin <- worldclim_global(var = "tmin", 
                         res = 5, 
                         path = "R/3_SDM/predictors/tmin") # 5 arc min ~ 10km ~ 0.08333 Dg
tmin <- stack(tmin)
# tmax
if(!dir.exists("R/3_SDM/predictors/tmax")){
  dir.create("R/3_SDM/predictors/tmax")
}
tmax <- worldclim_global(var = "tmax", 
                         res = 5, 
                         path = "R/3_SDM/predictors/tmax") # 5 arc min ~ 10km ~ 0.08333 Dg
tmax <- stack(tmax)
# temperature range
if(!dir.exists("R/3_SDM/predictors/trange")){
  dir.create("R/3_SDM/predictors/trange")
}
trange <- lapply(1:nlayers(stack(tmax)), function(x){
  tmin. <- raster(tmin[[x]])
  tmax. <- raster(tmax[[x]])
  return(abs(tmax.-tmin.))
})
trange <- stack(trange)
lapply(1:length(trange), function(x){
  raster::writeRaster(trange[[x]], 
                      paste0("R/3_SDM/predictors/trange/trange_",
                             formatC(x, width = 2, format = "d", flag = "0"),
                             ".tif"),
                      format = "GTiff",
                      overwrite=TRUE)
})
# prec
if(!dir.exists("R/3_SDM/predictors/prec")){
  dir.create("R/3_SDM/predictors/prec")
}
prec <- worldclim_global(var = "prec", 
                         res = 5, 
                         path = "R/3_SDM/predictors/prec") # 5 arc min ~ 10km ~ 0.08333 Dg
prec <- stack(prec)
# solar radiation
if(!dir.exists("R/3_SDM/predictors/solar")){
  dir.create("R/3_SDM/predictors/solar")
}
# calculate monthly solar radiation, defined for the year 2050, output to the current directory (1950 is the index year)
ETsolradRasters(rasterTemplate = raster(climatelayers[[1]]), 
                year = 0, # Start at 1950. So X is N years since 1950
                outputDir = "R/3_SDM/predictors/solar",
                overwrite=TRUE)
solar <- list.files("R/3_SDM/predictors/solar",full.names = T)
solar <- lapply(solar, raster)
solar <- stack(solar)
# calculate monthly PET
PET <- monthlyPET(tavg, solar, trange, tempScale = 1)
#--- Calculate aridity index ---#
# Precipitation (ppt) / Evapotranspiration (pet)
ppt <- raster::calc(prec,function(x) mean(x, na.omit=T))
pet <- raster::calc(PET,function(x) mean(x, na.omit=T))
Ari <- overlay(x = ppt, 
               y = pet, 
               fun = function(x, y){return(x / y)}) # Function to apply
Ari <- raster::crop(Ari, extent(-180,180,-60,90), snap='out')

precip <- prec
assignNames(precip = 'wc2.1_5m_prec_##')
thornwaite <- aridityIndexThornthwaite(precip, PET, precipScale = 1)

AI <- raster::crop(AI, extent(-180,180,-60,90), snap='out')
thornwaite <- raster::crop(thornwaite, extent(-180,180,-60,90), snap='out')

{
  plot(Ari)
  plot(AI)
  plot(thornwaite)
  plot(AI,Ari)
  plot(AI,thornwaite)
}

```

# Calculate ET0 - FAO Penman-Monteith equation described in 
## function
```{r}
ET0<-function(lat, z, TN, TX, RHmax, RHmin, FF, Rs, data, albedo=0.23, method_ea="allRH", method_ETo="PM") {
 
### Function to calculate Penman-Monteith and Hargreaves-Samani Potential Evapo-Transpiration, implementing the equations in Allen, R.G., L.S. Pereira, D. Raes, and M. Smith. 1998. ‘Crop evapotranspiration-Guidelines for Computing Crop Water requirements-FAO Irrigation and Drainage Paper 56’. FAO, Rome 300: 6541.
 
### Modified from Crespo, O, not dated (https://r-forge.r-project.org/scm/viewvc.php/*checkout*/convertBruceFormat/agriParameters.r?revision=280&root=xpos-r)
 
### For Penman-Monteith ETo, choice is given between the 4 methods described in Allen & al 1998 to compute the actual vapour pressure: using both RHmax and RHmin, RHmax only, RHmean or the minimum temperature to estimate dew temperature.
 
### required parameters:
 
## for all options:
 
# lat: latitude of the site in decimal degrees ([°])
 
# z: elevation above sea level [m]
 
# TN: daily minimum temperature at 2m height [°C]
 
# TX: daily maximum temperature at 2m height [°C]
 
# method_ETo: the equation used to compute ETo. Default is FAO Penman-Monteith (method_ETo="PM"). method_ETo="HS" gives the Hargreaves-Samani equation.
 
# data: a daily multivariate zoo time-series containing the required variables
 
## depending on the computation method chosen, see below:
 
# RHmax maximum relative humidity [%] and RHmin minimum relative humidity [%]
 
# or
 
# RHmax maximum relative humidity [%]
 
# or
 
# RHmean [%] = (RHmax+RHmin)/2
 
# Rs: incoming solar or shortwave radiation [MJ.m^(-2).day^(-1)]
 
# alb: albedo [dimensionless]. Default is alb=0.23 for the FAO hypothetical reference crop
 
# method_ea: method used to compute the actual vapour pressure in the Penman-Monteith equation. A choice is given between using both RHmax and RHmin (default, method_ea="allRH", RHmax only (method_ea="RHmax"), RHmean (method_ea="RHmean") or the minimum temperature to estimate dew temperature (method_ea=="Tdew").
 
##################
 
alb<-albedo
 
# Tmean
 
Tmean=(TN+TX)/2  ## Eq. 9 Allen & al. 1998.
 
# RHmean
 
if (method_ea=="RHmean") RHmean=(RHmax+RHmin)/2
 
#Slope of saturation vapour pressure curve
 
slopevap<-4098*0.6108*exp(17.27*Tmean/(Tmean+237.3))/((Tmean+237.3)^2) ## Eq. 13 Allen & al. 1998 [kPa/°C].
 
# calculation of psychometric constant
 
P <- 101.3*(((293-0.0065*z)/293)^5.26) ## Eq. 7 Allen & al. 1998 [kPa].
Cp <- 0.001013 ## Allen & al. 1998 p. 32.
epsilon <- 0.622 ## Allen & al. 1998 p. 32.
lambda <- 2.45 ## Allen & al. 1998 p. 31. O. Cresto uses lambda<-2.501-(0.002361*Tmean)
psychCon <- Cp*P/(epsilon*lambda) ## Eq. 8 Allen & al. 1998. but rounds values up.
 
# Mean saturation vapour pressure
eTmin <- 0.6108*exp(17.27*TN/(TN+237.3)) ## min saturation vapour pressure [kPa] Eq. 11 Allen & al. 1998.
eTmax <- 0.6108*exp(17.27*TX/(TX+237.3)) ## max saturation vapour pressure [kPa] Eq. 11 Allen & al. 1998.
es <- (eTmax+eTmin)/2 ## Eq. 12 Allen & al. 1998.
 
## Actual vapour pressure
if(method_ETo=="PM") {
# for RHmax and RHmin
if(method_ea=="allRH") ea<-(eTmin*RHmax+eTmax*RHmin)/200 ## Eq. 17 Allen & al. 1998.
# for RHmax
if(method_ea=="RHmax") ea<-eTmin*RHmax/100 ## Eq. 18 Allen & al. 1998
# for RH mean
if(method_ea=="RHmean") ea<-RHmean*(eTmax+eTmin)/200 ## Eq. 19 Allen & al. 1998
# In the absence of reliable relative humidity measurements
if(method_ea=="Tdew") ea<-0.6108*exp(17.27*TN/(TN+237.3)) ## Eq. 48 Allen & al. 1998, but see limitations p 58. Should be checked against measured
if(!method_ea %in% c("allRH", "RHmax", "RHmean", "Tdew")) print("Method for actual vapour pressure wrongly specified")
}
 
## Extraterrestrial radiation
# solar constant = 0.0820 MJ.m^(-2).min^(-1)
Gsc <- 0.0820
# convert site latitude to radians
phi <- pi*lat/180            # Eq. 23 Allen & al. 1998.
# julian day of the year
J <- as.numeric(format(index(data), "%j"))
# inverse relative distance Earth-Sun
Dr <- 1+0.033*cos(2*pi*J/365)      # Eq. 23 Allen & al. 1998.
# solar decimation [rad]
delta <- 0.409*sin((2*pi*J/365)-1.39) # Eq. 24 Allen & al. 1998.
# sunset hour angle [rad]
Ws <- acos(-tan(phi)*tan(delta))        # Eq. 25 Allen & al. 1998.
 
# Extraterrestrial radiation for daily periods [MJ.m^(-2).day^(-1)]
eRad <- (24*60/pi)*Gsc*Dr*(Ws*sin(phi)*sin(delta)+cos(phi)*cos(delta)*sin(Ws)) # Eq. 21 Allen & al. 1998.
## eRad <- zoo((24*60/pi)*Gsc*Dr*(Ws*sin(phi)*sin(delta)+cos(phi)*cos(delta)*sin(Ws)), date1)
 
if(method_ETo=="PM") {
# net solar radiation
Rns <- (1-albedo)*Rs ## Eq. 38 Allen & al. 1998. albedo = 0.23 for the FAO hypothetical crop reference
# clear-sky solar radiation
Rso <- (0.75+2*10^(-5)*z)*eRad ## Eq. 37 Allen & al. 1998
# net longwave radiation
SteBolCon <- 4.903*10^(-9) # Stefan-Boltzmann constant [4.903.10^(-9) MJ.K^(-4).m^(-2).day^(-1)]
Rnl <- SteBolCon*(((TX+273.16)^4+(TN+273.16)^4)/2)*(0.34 - 0.14*sqrt(ea))*((1.35*ifelse((Rs/Rso)>1,1,Rs/Rso))-0.35) ## Eq. 39 Allen & al. 1998.
# net radiation at the crop surface
Rn <- Rns-Rnl ## Eq. 40 Allen & al. 1998.
}
 
# Soil heat flux
G <-0 ## Eq. 42 Allen & al. 1998. G can be ignored for daily computations
 
# FAO Penman-Monteith equation for reference evapotranspiration
if(method_ETo=="PM") result <- (0.408*slopevap*(Rn-G)+(psychCon*900*FF*(es-ea)/(Tmean+273)))/(slopevap+psychCon*(1+0.34*FF))  ## Eq. 6 Allen & al. 1998.
# Hargreaves ETo
if(method_ETo=="HS") result <- 0.0023*(Tmean+17.8)*sqrt(TX-TN)*eRad/lambda ## Eq. 52 Allen & al. 1998; multiplied by 1/lambda to convert from MJ.m^(-2).day^(-1) to mm.day.day^(-1)
if(!method_ETo %in% c("PM", "HS")) print("ETo method wrongly specified")
 
return(result)
}
```

# Calc
```{r}
# get dataframe with vars I need

mydata <- stack(elev, tmin, tmax, )


ET <- 
  
```


## Soil
Soild variables from SoilGrids:
1. Depth to bedrock
2. pH
3. Clay proportion
4. Silt proportion
5. Bulk density

All soil variables were downloaded using Soildgrids 2.0 from the ftp server using the [geodata] package. Soildgrids 2.0 were way to large for downloading. There is no much difference between dataset at the resolution used here. Data on depth to bed rock is only available from Soilgrids version 1.0.
All soil variables were downloaded at a 2.5km resolution
Soils variables were included as it has been shown that climate change analyses that do not incorporate soils variability can misrepresent edaphic specialists (Corlett and Tomlinson 2020).

This map collection is described in the [META_GEOTIFF_1B.csv](https://files.isric.org/soilgrids/former/2017-03-10/data/META_GEOTIFF_1B.csv) spreadsheet.
Data is also downloadable from https://data.isric.org/

```{r soil}

if(!dir.exists("R/3_SDM/predictors/soil")){
  dir.create("R/3_SDM/predictors/soil")
}

# Soil data downloaded 09/28/21 using the ftp server

depths <- c(5, 15, 30, 60)    # the soil data is the average of the top 1 meter. 

cl <- makeCluster(length(depths))
clusterExport(cl, c("depths","soil_world","climatelayers"))
 
pblapply(depths,function(x){
  
  # 1. pH 
  soil_world(var = "phh2o", depth = x, stat="mean",
             path = "R/3_SDM/predictors/soil")
  # 2. Clay proportion 
  soil_world(var = "clay", depth = x, stat="mean",
             path = "R/3_SDM/predictors/soil")
  # 3. Silt proportion 
  soil_world(var = "silt", depth = x, stat="mean",
             path = "R/3_SDM/predictors/soil")
  # 4. Bulk density
  soil_world(var = "bdod", depth = x, stat="mean",
             path = "R/3_SDM/predictors/soil")
},
cl = cl)

stopCluster(cl)

# Download Depth to bedrock. Only available from soildGids 1.0
# Depth to bedrock
depthB <- "https://files.isric.org/soilgrids/former/2017-03-10/aggregated/5km/BDRICM_M_5km_ll.tif"
depthB <- raster(depthB)
depthB <- resample(depthB, raster(climatelayers[[1]]))
writeRaster(depthB, "R/3_SDM/predictors/depthB.tif",format = "GTiff", overwrite=TRUE)

##########################3
# 1) get average soil layers across the 4 depth measurements
# 2) resample to 5 arc min (~10km)
# 3) write raster

# pH
pH <- list.files("R/3_SDM/predictors/soil",pattern = "phh2o", full.names = T)
pH <- raster::stack(lapply(pH, raster::raster))
pH <- raster::calc(pH, fun = mean, na.rm = T)
pH <- raster::resample(pH, raster::raster(climatelayers[[1]]))
raster::writeRaster(pH, paste0("R/3_SDM/predictors/phh2o_avg.tif"), 
                    format = "GTiff")

# clay
clay <- list.files("R/3_SDM/predictors/soil",pattern = "clay", full.names = T)
clay <- raster::stack(lapply(clay, raster::raster))
clay <- raster::calc(clay, fun = mean, na.rm = T)
clay <- raster::resample(clay, raster::raster(climatelayers[[1]]))
raster::writeRaster(clay, paste0("R/3_SDM/predictors/clay_avg.tif"), 
                    format = "GTiff")

# silt
silt <- list.files("R/3_SDM/predictors/soil",pattern = "silt", full.names = T)
silt <- raster::stack(lapply(silt, raster::raster))
silt <- raster::calc(silt, fun = mean, na.rm = T)
silt <- raster::resample(silt, raster::raster(climatelayers[[1]]))
raster::writeRaster(silt, paste0("R/3_SDM/predictors/silt_avg.tif"), 
                    format = "GTiff")

# bulk
bulk <- list.files("R/3_SDM/predictors/soil",pattern = "bdod", full.names = T)
bulk <- raster::stack(lapply(bulk, raster::raster))
bulk <- raster::calc(bulk, fun = mean, na.rm = T)
bulk <- raster::resample(bulk, raster::raster(climatelayers[[1]]))
raster::writeRaster(bulk, paste0("R/3_SDM/predictors/bulk_avg.tif"), 
                    format = "GTiff")

```

# Load variables (if downloaded)
```{r save}


#################

varsIneed <- c("bio1","bio2","bio4","bio6","bio12","bio15")

# bioclimatic present
bios <- list.files("SDM/predictors/wc2.1_5m",
                   full.names = T,pattern = ".tif")
bios <- stack(lapply(bios, raster))
names(bios) <- gsub("wc2.1_5m_","",names(bios))
names(bios) <- gsub("_","",names(bios))
# Select climate variables 
varsIneed <- c("bio1","bio2","bio4","bio6","bio12","bio15")
bios <- bios[[varsIneed]]

# bioclimatic future

climatelayers_fut <- lapply(climatelayers_fut, function(x){
  x <- stack(x)
  names(x) <- paste0("bio",1:19)
  # Select climate variables 
  varsIneed <- c("bio1","bio2","bio4","bio6","bio12","bio15")
  x <- x[[varsIneed]]
  x
})

# Aridity and soil variables
asoil <- list.files("SDM/predictors",full.names = T,pattern = ".tif")
asoil <- stack(lapply(asoil, raster))

```

# Project to mollwide
```{r}
# present
bios <- projectRaster(bios, 
                      crs = "+proj=moll")

# future
climatelayers_fut <- lapply(climatelayers_fut, function(x){
  projectRaster(x, 
                crs = "+proj=moll")
})

# aridity and soil
asoil <- projectRaster(asoil, 
                       crs = "+proj=moll")

```

# Save predictors
```{r save}
if(!dir.exists("SDM/Cluster")){
  dir.create("SDM/Cluster")
  if(!dir.exists("SDM/Cluster/preds")){
    dir.create("SDM/Cluster/preds")
  }
}

# bioclimatic present
envs.names <- names(bios)

if(!dir.exists("SDM/Cluster/preds/bios_pres")){
  dir.create("SDM/Cluster/preds/bios_pres")
}

pblapply(1:nlayers(bios), function(x){
  raster::writeRaster(bios[[x]], 
                      paste0("SDM/Cluster/preds/bios_pres/",
                                        envs.names[x],".tif"), 
                    format = "GTiff",
                    overwrite=TRUE)
})

# Aridity and soil variables
envs.names <- names(asoil)

if(!dir.exists("SDM/Cluster/preds/soil")){
  dir.create("SDM/Cluster/preds/soil")
}

pblapply(1:nlayers(asoil), function(x){
  raster::writeRaster(asoil[[x]], 
                      paste0("SDM/Cluster/preds/soil/",
                                        envs.names[x],".tif"), 
                    format = "GTiff",
                    overwrite=TRUE)
})


#################
# future
if(!dir.exists("SDM/Cluster/preds/bios_fut")){
  dir.create("SDM/Cluster/preds/bios_fut")
}

for(i in 1:nrow(varstodown)){
  envs <- stack(climatelayers_fut[[i]])
  
  pblapply(1:nlayers(envs), function(x){
    raster::writeRaster(envs[[x]], 
                        paste0("SDM/Cluster/preds/bios_fut/",
                               varsIneed[x],"_",
                               varstodown[i,1],"_",
                               varstodown[i,2],
                               ".tif"), 
                        format = "GTiff",
                        overwrite=TRUE)
  })
}

```